---
title: "Recoding an NA and back again"
date: "12 March 2024"
date-modified: '12 March 2024'
categories:
  - R tips
author: Zoë Turner
subtitle: >
  "Replace NAs with `replace_na()` and code to NA with `na_if()`"
code-annotations: below
---

## Create data

```{r}
#| echo: false
#| include: false
library(tidyverse)
```

This was written originally in an Excel spreadsheet and used {datapasta} to copy into R as code to build the same data frame.
{datapasta} can be access through RStudio as an Addin as well as code.
Find out more about {datapasta} from the [Introduction to R and R Studio course](https://intro-r-rstudio.nhsrcommunity.com/session-datapasta.html#/title-slide).

## Recoding to NA

```{r}
survey <- tibble::tribble(
  ~Survey.Response, ~Code,
       "Response1",   -9L,
       "Response2",    2L,
       "Response3",   10L,
       "Response4",    0L,
       "Response5",    5L,
       "Response6",   -9L,
        "Missing", NA
  )

```

## Recode to NA

```{r}
library(tidyverse)

survey |> 
  mutate(new_column = na_if(Code, -9))
```

It's also possible to use the numbers and `case_when()`:

```{r}
survey |> 
  mutate(new_column = case_when(Code < 0 ~ NA,
                                Code == 0 ~ 1000,
                                .default = Code))

```

Or `ifelse()` where there are only two options:

```{r}
survey |> 
  mutate(new_column = ifelse(Code < 0, NA, Code))
```


## Recode from NA

```{r}
survey |> 
  mutate(new_column2 = replace_na(Code, 1000))
```
