---
title: "Showcasing a function - `separate()`"
date: "12 March 2024"
date-modified: '12 March 2024'
categories:
  - R tip
author: Zoë Turner
subtitle: >
  "Separating codes from one cell couldn't be any easier in R"
---

## Create data

```{r}
#| echo: false
#| include: false
library(tidyverse)
```

This was written originally in an Excel spreadsheet and used {datapasta} to copy into R as code to build the same data frame.
{datapasta} can be access through RStudio as an Addin as well as code.
Find out more about {datapasta} from the [Introduction to R and R Studio course](https://intro-r-rstudio.nhsrcommunity.com/session-datapasta.html#/title-slide).

```{r}
data <- tibble::tribble(
    ~Patient,          ~Codes,
  "PatientA", "A01, A02, A03",
  "PatientB", "B01; B02; B03",
  "PatientC", "C01; C03",
  "PatientD", "D01. D02. D03"
  )

```


## Separate codes by position

Separate into columns in the order data appears

```{r}
library(tidyverse)

data |> 
  tidyr::separate(Codes, c("col1", "col2", "col3"))
```


[https://tidyr.tidyverse.org/reference/separate.html](https://tidyr.tidyverse.org/reference/separate.html)

## Add a pivot

To move wide data to longer:

```{r}
data |> 
  tidyr::separate(Codes, c("col1", "col2", "col3")) |> 
  tidyr::pivot_longer(cols = c(starts_with("col")),
               names_to = "type")
```

