---
title: "Using {sf} to calculate catchment areas"
author:
  - name: Tom Jemmett
    orcid: 0000-0002-6943-2990
    email: thomas.jemmett@nhs.net
    affiliations:
      - name: The Strategy Unit
categories:
  - GIS
  - ggplot2
---

R is a highly capable [GIS][gis] tool, thanks to the `{sf}` package.
In this post I am going to show you how you can use `{sf}` with openly available
data to calculate catchment areas and population sizes, and then how to
calculate travel time isochrone’s with `{osrm}`. 

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(readxl)
  library(glue, include.only = "glue_data")
  library(scales, exclude = c("discard", "col_factor"))
  library(sf)
  library(leaflet)
  library(leaflet.extras)
  library(httr)
})
```

Firstly, I have prepared a dataset containing the 23 A&E departments that are
Adult Major Trauma Centres (MTCs) in England and Wales. This dataset is an
`{sf}` object which contains a special column called `geometry` - this column
contains the locations of the hospitals. The data is sourced from
[this][mts_map], the [Organisation Data Service][ods] (ODS) and the
[NHS Postcode Database][nhspd].

When we print this table out we get a bit of information telling us how many
features (rows) and fields (columns) there are, along with what the type of
the geometry is (in this case, we have points). We also get the "bounding box"
(bbox) - this is like a rectangle that encloses all of our geometries. Finally,
we get the [coordinate reference system][crs] (CRS) that is being used.

Below this we see a table of data - like any other dataframe in R.

```{r}
major_trauma_centres <- read_sf(
  paste0(
    "https://gist.githubusercontent.com/tomjemmett/",
    "ac111a045e1b0a60854d3d2c0fb0baa8/raw/",
    "a1a0fb359d1bc764353e8d55c9d804f47a95bfe4/",
    "major_trauma_centres.geojson"
  ),
  agr = "identity"
)
major_trauma_centres
```

This dataframe can be used like any other dataframe in R, for example, we can
use `{dplyr}` to filter rows out we aren't interested in, or join to another
dataframe to bring in extra information.

For example, we can load the LSOA population estimates and join it to the LSOA
boundaries file from the [ONS geoportal][lsoa_boundaries]. If you aren't
familiar with Output Areas see the [ONS Census Geography][oa] documentation.

First, let's load the population estimate file.

```{r}
#| message: false
#| warning: false

# download the LSOA population estimates from ons website: for some reason they
# provide this download as an excel file in a zip file, so we need to download
# the zip then extract the file, but we don't need to keep the zip after. withr
# handles this temporary like file for us
if (!file.exists("SAPE22DT2-mid-2019-lsoa-syoa-estimates-unformatted.xlsx")) {
  withr::local_file("lsoa_pop_est.zip", {
    download.file(
      paste0(
        "https://www.ons.gov.uk/file?uri=/peoplepopulationandcommunity/",
        "populationandmigration/populationestimates/datasets/",
        "lowersuperoutputareamidyearpopulationestimates/mid2019sape22dt2/",
        "sape22dt2mid2019lsoasyoaestimatesunformatted.zip"
      ),
      "lsoa_pop_est.zip",
      mode = "wb"
    )
    unzip("lsoa_pop_est.zip")
  })
}

lsoa_pop_estimates <- read_excel(
  "SAPE22DT2-mid-2019-lsoa-syoa-estimates-unformatted.xlsx",
  "Mid-2019 Persons",
  skip = 3
) %>%
  select(LSOA11CD = `LSOA Code`, pop = `All Ages`)

head(lsoa_pop_estimates)
```

Now we can load the boundary data and join to our population estimates.

```{r}
lsoa_boundaries <- read_sf(
  httr::modify_url(
    "https://services1.arcgis.com",
    path = c(
      "ESMARspQHYMw9BZ9",
      "arcgis",
      "rest",
      "services",
      "LSOA_Dec_2011_Boundaries_Generalised_Clipped_BGC_EW_V3",
      "FeatureServer",
      "0",
      "query"
    ),
    query = list(
      outFields="*",
      where="1=1",
      f="geojson"
    )
  )
) %>%
  # there are other fields in the lsoa data, but we only need the LSOA11CD field
  select(LSOA11CD) %>%
  inner_join(lsoa_pop_estimates, by = "LSOA11CD") %>%
  # this helps when combining different sf objects and gets rid of some messages
  st_set_agr(c(LSOA11CD = "identity", pop = "aggregate"))
```

We can now use ggplot to create a simple plot showing the populations of the
LSOAs and add points to show the MTCs.

```{r}
ggplot() +
  # first plot the lsoa boundaries and colour by the population
  geom_sf(data = lsoa_boundaries, aes(fill = pop), colour = NA) +
  geom_sf(data = major_trauma_centres) +
  scale_fill_distiller(type = "div", palette = "Spectral",
                       # tidy up labels in the legend, display as thousands
                       labels = number_format(accuracy = 1e-1, scale = 1e-3, suffix = "k")) +
  theme_void() +
  theme(legend.position = c(0, 0.98),
        legend.justification = c(0, 1)) +
  labs(fill = "Population (thousands)")
```

We can now try to calculate the population estimates for each of the MTCs. We
can take advantage of {sf} to perform a spatial join between our two datasets.
A spatial join is similar to a join in a regular SQL database, except instead of
joining on columns we join on geospatial features.

In this case we want to take each LSOA and find which MTC it is closest to, so
we can use the `st_nearest_feature` predicate function with `st_join`. This will
give us a dataset containing all of the rows from `lsoa_boundaries`, but
augmented with the columns from `major_trauma_centres`.

We can then use standard `{dplyr}` functions to group by the MTC `name` and
`org_id` fields, and calculate the sum of the `pop` column.

`{sf}` is clever - it knows that when we are summarising a geospatial table we
need to combine the geometries somehow. What it will do is call `st_union` and
return us a single geometry per MTC.

One thing we need to do before joining our data though is to transform our data
temporarily to a different CRS. Our data currently is using latitude and
longitude in a spherical projection, but `st_nearest_points` assumes that the
points are planar. That is, it assumes that the world is flat. This can lead to
incorrect results.

But, we can transform the data to use the [British National Grid][bng]. This
instead specifies points as how far east and north from an origin. This has a
CRS number of 27700. Once we are done summarising our data we can again
project back to the previous CRS (4326). [This][understanding_crs] article
explains these numbers in a bit more detail.

```{r}
mtcv_pop <- lsoa_boundaries %>%
  st_transform(crs = 27700) %>%
  # assign each LSOA to a single, closest, MTC
  st_join(
    st_transform(major_trauma_centres, crs = 27700),
    join = st_nearest_feature
  ) %>%
  # now, summarise our data frame based on MTCs to find the total population
  group_by(name, org_id) %>%
  # note: summarise will automatically call st_union on the geometries
  # this gives us the lsoas as a combined multipolygon
  summarise(across(pop, sum), .groups = "drop") %>%
  st_transform(crs = 4326)
```

We can now visualise this map. This time we will use the `{leaflet}` package to
create an interactive html map.

First we need a function for colouring the areas.

```{r}
pal_fun <- colorNumeric("Spectral", NULL, n = 7, reverse = TRUE)
```

As this is an interactive map we can add popups to the areas to make it easier
to identify which MTC we are looking at and to show us the estimated population.

```{r}
p_popup <- glue_data(mtcv_pop,
                     "<strong>{name}</strong> ({org_id})",
                     "Estimated Population: {comma(pop)}",
                     .sep = "\n")
```

Now we can create our map

```{r}
mtcv_pop %>%
  leaflet() %>%
  # add the areas that have been assigned to each MTC
  addPolygons(stroke = TRUE,
              color = "black",
              opacity = 1,
              weight = 1,
              fillColor = ~pal_fun(pop),
              fillOpacity = 1,
              popup = p_popup) %>%
  # add markers for the MTCs
  addCircleMarkers(data = major_trauma_centres,
                   color = "black",
                   stroke = TRUE,
                   fillColor = "white",
                   weight = 1,
                   radius = 3,
                   opacity = 1,
                   fillOpacity = 1) %>%
  setMapWidgetStyle(list(background= "white"))
```

This method certainly is not perfect. For example, we can see that the closest
MTC for a large part of Somerset is Cardiff, but in order to reach Cardiff you
need to cross the Bristol Channel! Not so easy by a Land Ambulance!

A better approach would be to use some method to calculate travel time from each
LSOA to the MTCs, an example of using the Google Maps Travel Time API is
covered [here][gapi_travel]. However, as a quick and dirty approach this method
doesn't produce too bad results.

[gis]: https://en.wikipedia.org/wiki/Geographic_information_system
[mts_map]: https://www.nhs.uk/NHSEngland/AboutNHSservices/Emergencyandurgentcareservices/Documents/2016/MTS-map.pdf
[ods]: https://digital.nhs.uk/services/organisation-data-service
[nhspd]: https://geoportal.statistics.gov.uk/datasets/nhs-postcode-directory-uk-full-november-2020
[crs]: https://en.wikipedia.org/wiki/Spatial_reference_system
[lsoa_boundaries]: https://geoportal.statistics.gov.uk/datasets/lower-layer-super-output-areas-december-2011-boundaries-super-generalised-clipped-bsc-ew-v3?geometry=-25.530%2C48.013%2C21.184%2C57.298
[oa]: https://www.ons.gov.uk/methodology/geography/ukgeographies/censusgeography#output-area-oa
[bng]: https://en.wikipedia.org/wiki/Ordnance_Survey_National_Grid
[understanding_crs]: https://www.earthdatascience.org/courses/earth-analytics/spatial-data-r/understand-epsg-wkt-and-other-crs-definition-file-types/
[gapi_travel]: https://hwangnyc.medium.com/calculating-travel-time-and-distance-using-google-maps-api-in-r-bbc5b74df066
