---
title: "Using R to create column charts featuring 95% confidence intervals"
date: "5 January 2022"
date-modified: "27 July 2024"
categories:
  - Statistics
  - tidyverse
  - ggplot2
  - Public Health
author: Daniel Weiand
subtitle: >
---

### **Daniel Weiand – Consultant Medical Microbiologist – Newcastle upon Tyne NHS Foundation Trust**

Hello!

This is my first blog post for the NHS R Community, which I stumbled across in the course of my work as a consultant medical microbiologist at Newcastle upon Tyne Hospitals NHS Foundation Trust.

At work, I’ve been trying to use R to create column charts featuring 95% confidence intervals. I approached the friendly people on the NHS R Community’s Slack channel for further information and guidance.

I must add here that the Community’s Slack channel has been extremely helpful to me, as a novice R user, in solving some of the issues I’ve experienced, and highlighting R packages of potential interest. This is the first time I’ve tried to create a ReprEx and now I understand why people love (?) the mtcars database as much as they do!

### **Step 1: Calculate some summary statistics**

I wanted to calculate some summary statistics, including the mean, and standard error or 95% confidence intervals.

Initially I came across the `summary()` function of Base R, which is helpful as it calculates the `Min.`, `1st Qu.`, `Median`, `Mean`, `3rd Qu.`, and `Max`.

However, the `summary()` function of {base} R does not calculate either the standard error or the 95% confidence intervals:

```{r}
library(tidyverse)
# calculate summary statistics for all numeric data using summary() and where(is.numeric())
mtcars %>%
  select(where(is.numeric)) %>%
  summary()

# calculate summary statistics for mpg using summary() and where(is.numberic())
mtcars %>%
  select(mpg) %>%
  summary()
```

Then zx8754 very kindly pointed me towards a method for calculating the standard error on StackOverflow: <https://stackoverflow.com/q/2676554/680068>

```{r}
# create stderr function
stderr <- function(x, na.rm = TRUE) {
 if (na.rm) x <- na.omit(x)
 sqrt(var(x)/length(x))
}
```

Then I used this function to calculate summary statistics, incl. mean and standard error, using the `summarise()` and `across()` functions of {dplyr}:

```{r}
# calculate summary statistics using summarise() and across() and n/mean/min/median/max/sd/stderr
# stderr <- function(x, na.rm=TRUE) {
#  if (na.rm) x <- na.omit(x)
#  sqrt(var(x)/length(x))
# }
mtcars %>% 
 group_by(cyl) %>% 
 mutate(
   across(mpg, 
          list(
            n = ~ n(),
            mean = ~ mean(.x, na.rm = TRUE),
            min = ~ min(.x, na.rm = TRUE),
            median = ~ median(.x, na.rm = TRUE),
            max = ~ max(.x, na.rm = TRUE),
            sd = ~ sd(.x, na.rm = TRUE),
            stderr = ~ stderr(.x)),
          .names = NULL)) %>%
 select(starts_with("mpg")) %>% 
 summarise(mean = mean(mpg_mean),
           min = mean(mpg_min),
           median = mean(mpg_median),
           max = mean(mpg_max),
           sd = mean(mpg_sd),
           stderr = mean(mpg_stderr)) %>% 
#create column chart with error bars (using stderr) 
 ggplot(aes(cyl, mean))+
 geom_col(na.rm = TRUE)+
 geom_errorbar(aes(ymin = mean-stderr, ymax = mean + stderr), 
               position = "dodge", width = 0.25)
#calculate summary statistics using summarise() and across() and n/mean/min/median/max/sd/stderr
# stderr <- function(x, na.rm=TRUE) {
#  if (na.rm) x <- na.omit(x)
#  sqrt(var(x)/length(x))
# }
mtcars %>% 
 group_by(cyl) %>% 
 mutate(
   across(mpg, 
          list(
            n = ~ n(),
            mean = ~ mean(.x, na.rm = TRUE),
            min = ~ min(.x, na.rm = TRUE),
            median = ~ median(.x, na.rm = TRUE),
            max = ~ max(.x, na.rm = TRUE),
            sd = ~ sd(.x, na.rm = TRUE),
            stderr = ~ stderr(.x)),
          .names = NULL)) %>%
 select(starts_with("mpg")) %>% 
 summarise(mean = mean(mpg_mean),
           min = mean(mpg_min),
           median = mean(mpg_median),
           max = mean(mpg_max),
           sd = mean(mpg_sd),
           stderr = mean(mpg_stderr)) %>% 
#create column chart with error bars (using stderr) 
 ggplot(aes(cyl, mean))+
 geom_col(na.rm = TRUE)+
 geom_errorbar(aes(ymin = mean-stderr, ymax = mean + stderr), 
               position = "dodge", width = 0.25)

```

### **Step 2: Create column charts with error bars (using 95% confidence intervals)**

Then Seb Fox pointed me towards a method for calculating 95% confidence intervals using the {PHEindicatormethods} package, available on CRAN: <https://cran.r-project.org/web/packages/PHEindicatormethods/index.html>

```{r}
# create MEAN column chart with error bars (using 95% confidence intervals) 
require(PHEindicatormethods)

mtcars %>% 
 filter(!is.na(cyl)) %>% 
 group_by(cyl) %>% 
#use phe_mean()
 phe_mean(x = mpg, # field name from data containing the values to calculate the means for
          type = "full", # defines the data and metadata columns to be included in output; can be "value", "lower", "upper", "standard" (for all data) or "full" (for all data and metadata); quoted string; default = "full"
          confidence = 0.95) %>% #required level of confidence expressed as a number between 0.9 and 1
  
# create column chart with error bars (using 95% CI calculated using phe_mean())
 ggplot(aes(cyl, value))+
 geom_col(na.rm = TRUE)+
 geom_errorbar(aes(ymin = lowercl, ymax = uppercl), 
               position = "dodge", width = 0.25)
 
# create PROPORTION column chart with error bars (using 95% confidence intervals) 
require(PHEindicatormethods)

mtcars %>% 
 group_by(cyl) %>% 
 summarise(n = n(),
           sum = sum(n)) %>% 
 mutate(sum = sum(n)) %>%
# phe_proportion()
 phe_proportion(x = n, # numerator
                n = sum, # denominator
                type = "full", # defines the data and metadata columns to be included in output; can be "value", "lower", "upper", "standard" (for all data) or "full" (for all data and metadata); quoted string; default = "full"
                confidence = 0.95, # required level of confidence expressed as a number between 0.9 and 1 
                multiplier = 100) %>%   # the multiplier used to express the final values (eg 100 = percentage); numeric; default 1
# create column chart with error bars (using 95% CI calculated using phe_proportion())
 ggplot(aes(cyl, value))+
 geom_col(na.rm = TRUE)+
 geom_errorbar(aes(ymin = lowercl, ymax = uppercl), 
               position = "dodge", width = 0.25)

```

I hope that the code, above, helps a few colleagues of mine across the NHS, in some small way.

Thank you, again, to all members of the NHS R Community, for all your help. Particular thanks go to everyone who has helped me, to date, on the NHS R Community’s Slack channel.

### **Comments (3)**

1.  **Chuck Burks**

    5 January 2022

    I used to use a function like that, then I realized that I could get a function for standard error through the FSA package, `FSA::se()`. Arguments can be made either way; why reinvent the wheel vs. why load a large package to avoid writing one function.

2.  **Qin Zeng**

    5 January 2022

    This works out about the same: `mtcars %>% group_by(cyl) %>% summarise_at(vars(mpg), funs(n(), mean, min, median, max, sd, stderr)) %>% ggplot(aes(cyl, mean))+ geom_col(na.rm = TRUE)+ geom_errorbar(aes(ymin = mean-stderr, ymax = mean+stderr), position = "dodge", width = 0.25)`

3.  **Stephen**

    6 January 2022

    Two suggestions: `mtcars %>% group_by(cyl) %>% summarise(n = n(), sum = sum(n)) %>% mutate(sum = sum(n))` is more easily said as `mtcars %>% count(cyl) %>% mutate(sum = sum(n))` Qin, your sequence is better as mtcars %\>% group_by(cyl) %\>% summarise(n(), across(mpg, c(mean, min, median, max, sd, stdeQi, in modern dplyr (summarise_at is deprecated)
